Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-06-18 manuscript/manuscript.Rmd
add section 2.1

As a simple illustration, Figure 1 shows how comparing means between two conditions conceal the shapes of the underlying RT and accuracy distributions. We simulated a RT + accuracy data set for a single subject who performed 200 trials (i.e., repeated measurements) in each of two conditions. For example, while this subject is 71 ms faster on average in condition 1 (481 ms) compared to condition 2 (552 ms), the corresponding hazard functions of response occurrence show that the effect starts in time period (400,500] or bin t=5, and is present in three consecutive time bins (i.e., for 300 ms). Similarly, while this subject makes less errors in condition 1 (86% accuracy) compared to condition 2 (64% accuracy), the conditional accuracy functions show that (a) erroneous responses in condition 1 are confined to a single time bin, and (b) the observed conditional accuracies (0, 1, 0.51, 0.48) are never even close to the mean accuracies.

@@

(ref:descr-fig1-caption) Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

>>>>

As a simple illustration, Figure 1 shows how comparing means between two conditions conceal the shapes of the underlying RT and accuracy distributions. We simulated a RT + accuracy data set for a single subject who performed 200 trials (i.e., repeated measurements) in each of two conditions. For example, while this subject is 71 ms faster on average in condition 1 (481 ms) compared to condition 2 (552 ms), the corresponding hazard functions of response occurrence show that the effect starts in time period (400,500] or bin t=5, and is present in three consecutive time bins (i.e., for 300 ms). Similarly, while this subject makes less errors in condition 1 (86% accuracy) compared to condition 2 (64% accuracy), the conditional accuracy functions show that (a) the effect is present only for responses emitted before 400 ms, (b) erroneous responses in condition 1 are confined to a single time bin, and (c) the observed conditional accuracies (0, 1, 0.51, 0.48) are never even close to the mean accuracies.

@@

(ref:descr-fig1-caption) Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B). Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (< 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (>500 ms) are controlled by the target stimulus.

@@

## 2.1 Single, repeatable, and recurrent events

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus that (s)he has to detect by pressing a button within a certain time window (e.g., 1 second from target onset). The single time-to-detection is thus repeatedly measured across multiple trials. In the parlance of EHA, in each trial the participant is "at risk" for response occurrence as long as the response has not occurred yet.

In a two-button discrimination task, the participant is presented in each trial with a target stimulus that (s)he has to categorize by pressing one of two buttons within a certain time window. This is known as a "competing risks" situation, because in each trial, the participant can transition from an idle state to either a correct response state or an incorrect response state.

In a bistable perception task, the participant is looking at an ambiguous stimulus (e.g., the duck-rabbit illusion, the Necker cube) for two minutes, for example, and asked to press a button each time when her/his perception switches from one possibe interpretation to the other possible interpretation. In this task, there are two events (percept A switches to percept B, percept B switches to percept A) that can recur within the same observation period (2 minutes). We do not analyse recurrent events in this tutorial. More information about recurrent events analysis can be found in REF and REF...

In secion A of the Supplementary Material we visualize the types of time-to-event data that can be obtained in these RT tasks.

## 2.2 Right censoring vs data trimming

## 2.3 Discrete vs continuous time units

All measurements of duration are discrete in nature. However, when the temporal resolution is high relative to the duration of the observation window, 
- interest in the shape of the hazard function <-> Cox
- non-smooth functions <-> parametric
- definition hazard prob and logistic regression

## 2.4 Discrete-time Hazard functions and conditional accuracy fun

def hazard

def survival

def cumulative distribution

def prob mass function

def conditional accuracz function

## 2.5 Number of time bins, repeated measures, and samples

## 2.6 Bayesian vs. frequentist approach

-------------------

### 2025-06-18 manuscript/manuscript.tex
add section 2.1

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the early emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higer in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (\textgreater500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1.

For many psychological questions, such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, as well as adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit. Thus, a distributional approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to better discriminate between different theoretical accounts of psychological and/or brain-based processes.

@@

>>>>

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higer in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (\textgreater500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1.

For many psychological questions, such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, by adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit. Thus, a distributional approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to better discriminate between different theoretical accounts of psychological and/or brain-based processes.

@@

@@

\subsection{2.1 Single, repeatable, and recurrent events}\label{single-repeatable-and-recurrent-events}

While people can die only once, in experimental RT tasks the events of interest - button presses, saccade onsets, etc. - are typically repeatable. For example, in a one-button detection task the participant is presented in each trial with a faint target stimulus that (s)he has to detect by pressing a button within a certain time (e.g., 1 second).

Detection: single event, repeatable: hazard function + clustering
Discrimination: Two events, repeatable: competing risks
Bistable perception task: Two events, repeatable, and recurrent

\subsection{2.2 Right censoring vs data trimming}\label{right-censoring-vs-data-trimming}

\subsection{2.3 Discrete vs continuous time units}\label{discrete-vs-continuous-time-units}

All measurements of duration are discrete in nature. However, when the temporal resolution is high relative to the duration of the observation window,
- interest in the shape of the hazard function \textless-\textgreater{} Cox
- non-smooth functions \textless-\textgreater{} parametric
- definition hazard prob and logistic regression

\subsection{2.4 Discrete-time Hazard functions vs cumulative distribution functions}\label{discrete-time-hazard-functions-vs-cumulative-distribution-functions}

def hazard

def survival

def cumulative distribution

def prob mass function

def conditional accuracz function

\subsection{2.5 Number of time bins, repeated measures, and samples}\label{number-of-time-bins-repeated-measures-and-samples}

\subsection{2.6 Bayesian vs.~frequentist approach}\label{bayesian-vs.-frequentist-approach}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
